---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
1. Load the libraries
```{r libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(lattice)
library(timeDate)
```

2. Load the data
```{r data, echo=TRUE}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
```{r total steps per day, echo=TRUE}
total_steps_per_day <- activity[,c(1,2)] %>% 
  group_by(date) %>%
  summarise(steps = sum(steps, na.rm = TRUE)) %>%
  as.data.frame()
head(total_steps_per_day)
```

2. Make a histogram of the total number of steps taken each day
```{r histogram of total steps per day, echo=TRUE}
qplot(total_steps_per_day[,2], 
      geom = "histogram",
      main = "Histogram of total number of steps taken each day",
      xlab = "steps")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean and median of total steps per day, echo=TRUE}
mean(total_steps_per_day[,2])
median(total_steps_per_day[,2])
```

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days 
```{r time series plot of average steps a day, echo=TRUE}
average_steps_a_day <- activity %>% 
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE)) %>%
  as.data.frame()

plot(average_steps_a_day[,1],
     average_steps_a_day[,2],
     type = "l",
     main = " Average number of steps taken averaged across all days",
     xlab = "interval(min)",
     ylab = "steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
```{r interval with max average steps a day, echo=TRUE}
average_steps_a_day[order(average_steps_a_day$steps, decreasing = TRUE),][1,1]
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)
```{r total missing values in the dataset, echo=TRUE}
sum(is.na(activity[,1]))
```

2. Devise a strategy for filling in all of the missing values in the dataset.
The strategy does not need to be sophisticated. For example, you could use 
the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r filling in all of the missing values, echo=TRUE}
temp <- activity
temp[is.na(activity),1] <- mean(temp[,1], na.rm = TRUE)
head(temp)
```

3. Create a new dataset that is equal to the original dataset
but with the missing data filled in.
```{r missing values filled dataset, echo=TRUE}
filled_activity <- temp
```

4. Make a histogram of the total number of steps taken each day and 
Calculate and report the mean and median total number of steps taken
per day. Do these values differ from the estimates from the first part 
of the assignment? What is the impact of imputing missing data on 
the estimates of the total daily number of steps?
```{r plot mean and median of filled dataset, echo=TRUE}
filled_total_steps_per_day <- filled_activity[,c(1,2)] %>% 
  group_by(date) %>%
  summarise(steps = sum(steps)) %>%
  as.data.frame()

qplot(total_steps_per_day[,2], 
      geom = "histogram",
      main = "Histogram of total number of steps taken each day",
      xlab = "steps")

qplot(filled_total_steps_per_day[,2], 
      geom = "histogram",
      main = "Histogram of total number of steps taken each day with NA overwritten",
      xlab = "steps")

mean(filled_total_steps_per_day[,2])
median(filled_total_steps_per_day[,2])
```

## Are there differences in activity patterns between weekdays and weekends?
1.Create a new factor variable in the dataset with two levels "weekday" 
and "weekend" indicating whether a given date is a weekday or weekend day
```{r weekday marked activity, echo=TRUE, warning=FALSE}
weekday_logical <- isWeekday(as.Date(activity[,2]))
weekday_activity <- mutate(activity, 
                           weekday = factor(weekday_logical, 
                                            levels = c(TRUE, FALSE), 
                                            labels = c("weekday", "weekend")))
head(weekday_activity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis). 
```{r panel plot, echo=TRUE}
weekday_average_steps_a_day <- weekday_activity %>% 
  group_by(interval, weekday) %>%
  summarise(steps = mean(steps, na.rm = TRUE)) %>%
  as.data.frame()

xyplot(steps~interval | weekday, 
       data = weekday_average_steps_a_day, 
       type = "l", 
       layout = c(1,2),
       main = "Average number of steps, averaged across 
       all weekday days or weekend days ")
```